### 2023-10-30 journal/analysis.tex
created directory for eventual journal version
Updates to related work--incorporating new info about Jayanti et al.'s lower bound
editing Section 3; also had to get rid of cleverref since it is incompatible with journal style

\begin{lemma}\label{dSearchTime}
The search that \op{FindResponse}$(b,i)$ does at line \ref{FRb} to find the index $b_e$ of the block in the root containing the $e$th enqueue takes $O(\log (\var{root}.\fld{blocks}[b_e].\size + \var{root}.\fld{blocks}[b-1].\size))$ steps.
\end{lemma}
\begin{proof}
Let the blocks in the root be $B_1, \ldots, B_\ell$.
The doubling search for $b_e$ takes $O(\log (b-b_e))$ steps,
so we prove $b - b_e \leq 2 \cdot B_{b_e}.\size + B_{b-1}.\size + 1$.
If $b \leq b_e+1$, then this is trivial, so assume  $b>b_e+1$.
%
As shown in \Cref{linearCorrect}, the dequeue that calls \op{FindResponse} is in $B_b$ and is supposed to return an enqueue in $B_{b_e}$.
Thus, there can be at most $B_{b_e}.\size$ dequeues in 
$D(B_{b_e+1}) \cdots D(B_{b-1})$; otherwise in the sequential execution $L$,
all elements enqueued before the end of
$E(B_{b_e})$ would be dequeued before $D(B_b)$. 
Furthermore, by \Cref{sizeCorrectness}, the size of the queue  after the prefix of $L$ corresponding to 
$B_1,\ldots,B_{b-1}$  is 
$B_{b-1}.\size \geq B_{b_e}.size + |E(B_{b_e+1})\cdots E(B_{b-1})| - |D(B_{b_e+1})\cdots D(B_{b-1})|$.
Thus, $|E(B_{b_e+1})\cdots E(B_{b-1})| \leq B_{b-1}.\size + |D(B_{b_e+1})\cdots D(B_{b-1})| \leq B_{b-1}.\size + B_{b_e}.\size$.
So, the total number of operations in $B_{b_e+1}, \ldots, B_{b-1}$ is at most
$B_{b-1}.\size + 2\cdot B_{b_e}.\size$.
Each of these $b-1-b_e$ blocks contains at least one operation, by \Cref{blockNotEmpty}.
So, $b-1-b_e \leq B_{b-1}.\size + 2\cdot B_{b_e}.\size$.
\end{proof}

The following lemma helps bound the time for \op{GetEnqueue}.
\begin{lemma}\label{blockSize}
Each block $B$ in each node contains at most one operation of each process.
If $c$ is the execution's maximum point contention, $B$ has at most $c$ direct~subblocks.
\end{lemma}
\begin{proof}
Suppose $B$ contains an operation of process $p$.
Let $op$ be the earliest operation by $p$ contained in $B$.
When $op$ terminates, $op$ is contained in $B$ by \Cref{lem::appendExactlyOnce}.
Thus, $B$ cannot contain any later operations by $p$, since $B$ is created before
those operations are invoked.

Let $t$ be the earliest termination of any operation contained in $B$.
By \Cref{lem::appendExactlyOnce}, $B$ is created before $t$, so all operations contained in $B$
are invoked before $t$.  Thus, all are  running concurrently at $t$, so $B$ contains at most $c$ operations.
By definition, the direct subblocks of $B$ contain these $c$ operations, and each operation is contained
in exactly one of these subblocks, by Lemma \ref{lem::subblocksDistinct}.
By \Cref{blockNotEmpty}, each direct subblock of $B$ contains at least one operation,
so $B$ has at most $c$ direct subblocks.
\end{proof}

@@

A non-null \op{Dequeue} must also search at line \ref{FRb} and call \op{GetEnqueue}
at line \ref{findAnswer}.
By \Cref{dSearchTime}, the doubling search takes $O(\log(q_e+q_d+p))$ steps,
since the size of the queue can change by at most $p$ within one block (by \Cref{blockSize}).
\op{GetEnqueue} does a binary search within each node on a path from the root to a leaf.
Each node $\var{v}$'s search is within the subblocks of one block in $\var{v}$'s parent.
By \Cref{blockSize}, each such search takes $O(\log c)$ steps, for a total of $O(\log p\log c)$ steps.
\end{proof}

>>>>

\begin{lemma}\label{dSearchTime}
The search that \op{FindResponse}$(b,i)$ does at line \ref{FRb} to find the index $b_e$ of the block in the root containing the $e$th enqueue takes $O(\log (\var{root}.\fld{blocks}[b_e].\size + \var{root}.\fld{blocks}[b-1].\size))$ steps.
\end{lemma}
\begin{proof}
Let the blocks in the root be $B_1, \ldots, B_\ell$.
The doubling search for $b_e$ takes $O(\log (b-b_e))$ steps,
so we prove $b - b_e \leq 2 \cdot B_{b_e}.\size + B_{b-1}.\size + 1$.
If $b \leq b_e+1$, then this is trivial, so assume  $b>b_e+1$.
%
As shown in Lemma \ref{linearCorrect}, the dequeue that calls \op{FindResponse} is in $B_b$ and is supposed to return an enqueue in $B_{b_e}$.
Thus, there can be at most $B_{b_e}.\size$ dequeues in 
$D(B_{b_e+1}) \cdots D(B_{b-1})$; otherwise in the sequential execution $L$,
all elements enqueued before the end of
$E(B_{b_e})$ would be dequeued before $D(B_b)$. 
Furthermore, by Lemma \ref{sizeCorrectness}, the size of the queue  after the prefix of $L$ corresponding to 
$B_1,\ldots,B_{b-1}$  is 
$B_{b-1}.\size \geq B_{b_e}.size + |E(B_{b_e+1})\cdots E(B_{b-1})| - |D(B_{b_e+1})\cdots D(B_{b-1})|$.
Thus, $|E(B_{b_e+1})\cdots E(B_{b-1})| \leq B_{b-1}.\size + |D(B_{b_e+1})\cdots D(B_{b-1})| \leq B_{b-1}.\size + B_{b_e}.\size$.
So, the total number of operations in $B_{b_e+1}, \ldots, B_{b-1}$ is at most
$B_{b-1}.\size + 2\cdot B_{b_e}.\size$.
Each of these $b-1-b_e$ blocks contains at least one operation, by Corollary \ref{blockNotEmpty}.
So, $b-1-b_e \leq B_{b-1}.\size + 2\cdot B_{b_e}.\size$.
\end{proof}

The following lemma helps bound the time for \op{GetEnqueue}.
\begin{lemma}\label{blockSize}
Each block $B$ in each node contains at most one operation of each process.
If $c$ is the execution's maximum point contention, $B$ has at most $c$ direct~subblocks.
\end{lemma}
\begin{proof}
Suppose $B$ contains an operation of process $p$.
Let $op$ be the earliest operation by $p$ contained in $B$.
When $op$ terminates, $op$ is contained in $B$ by Lemma \ref{lem::appendExactlyOnce}.
Since $B$ is created before
those operations are invoked, $B$ cannot contain any later operations by $p$.

Let $t$ be the earliest termination of any operation contained in $B$.
By Lemma \ref{lem::appendExactlyOnce}, $B$ is created before $t$, so all operations contained in $B$
are invoked before $t$.  Thus, all are  running concurrently at $t$, so $B$ contains at most $c$ operations.
By definition, the direct subblocks of $B$ contain these $c$ operations, and each operation is contained
in exactly one of these subblocks, by Lemma \ref{lem::subblocksDistinct}.
By Corollary \ref{blockNotEmpty}, each direct subblock of $B$ contains at least one operation,
so $B$ has at most $c$ direct subblocks.
\end{proof}

@@

A non-null \op{Dequeue} must also search at line \ref{FRb} and call \op{GetEnqueue}
at line \ref{findAnswer}.
By Lemma \ref{dSearchTime}, the doubling search takes $O(\log(q_e+q_d+p))$ steps,
since the size of the queue can change by at most $p$ within one block (by Lemma \ref{blockSize}).
\op{GetEnqueue} does a binary search within each node on a path from the root to a leaf.
Each node $\var{v}$'s search is within the subblocks of one block in $\var{v}$'s parent.
By Lemma \ref{blockSize}, each such search takes $O(\log c)$ steps, for a total of $O(\log p\log c)$ steps.
\end{proof}

### 2023-10-30 journal/boundedspace.tex
created directory for eventual journal version
Updates to related work--incorporating new info about Jayanti et al.'s lower bound
reformatting GC code
editing Section 3; also had to get rid of cleverref since it is incompatible with journal style

Operations remain in the \fld{blocks} arrays forever. 
This uses space proportional to the number of enqueues that have been invoked.
Now, we modify the implementation to remove blocks that are no longer needed, so that space usage is
polynomial in $p$ and $q$, while (amortized) step complexity is still polylogarithmic.  For lack of space, details are in Appendix~\ref{reducing-details}.
We replace the \fld{blocks} array in each node by a red-black tree (RBT)
that stores the blocks.
Each block has an additional \fld{index} field
that represents its position within the original \fld{blocks} array, and
blocks in a RBT are sorted by \fld{index}.
The attempt to install a new block in $\fld{blocks}[i]$  on line \ref{cas}
is replaced by an attempt to insert a new block with index $i$ into the RBT.
Accessing the block in $\fld{blocks}[i]$ is 
replaced by searching the RBT for the  index $i$.
The binary searches for a block in line \ref{FRb} and \ref{getChild} can simply search the RBT
using the \fld{sum\sub{enq}} field, since the RBT is also sorted with respect to this field, by \Cref{lem::sum}.

>>>>

Operations remain in the \fld{blocks} arrays forever. 
This uses space proportional to the number of enqueues that have been invoked.
Now, we modify the implementation to remove blocks that are no longer needed, so that space usage is
polynomial in $p$ and $q$, while (amortized) step complexity is still polylogarithmic. We replace the \fld{blocks} array in each node by a red-black tree (RBT)
that stores the blocks.
Each block has an additional \fld{index} field
that represents its position within the original \fld{blocks} array, and
blocks in a RBT are sorted by \fld{index}.
The attempt to install a new block in $\fld{blocks}[i]$  on line \ref{cas}
is replaced by an attempt to insert a new block with index $i$ into the RBT.
Accessing the block in $\fld{blocks}[i]$ is 
replaced by searching the RBT for the  index $i$.
The binary searches for a block in line \ref{FRb} and \ref{getChild} can simply search the RBT
using the \fld{sum\sub{enq}} field, since the RBT is also sorted with respect to this field, by Invariant \ref{inv::sum}.

### 2023-10-30 journal/conclusion.tex
editing Section 3; also had to get rid of cleverref since it is incompatible with journal style

We believe the approach used here to implement a lock-free queue 
could be applied to obtain other lock-free
data structures with a polylogarithmic step complexity.
For example, we can easily adapt our routines to implement a  vector data structure that stores a sequence and
provides three operations: \opa{Append}{e} to add an element \var{e} to the end of the sequence,
\opa{Get}{i} to read the \var{i}th element in the sequence, and
\opa{Index}{e} to compute the position of element \var{e} in the sequence.
%\here{Can omit rest of this paragraph to save space}
%An \opa{append}{e} is implemented like \opa{Enqueue}{e} in $O(\log p)$ steps.  
%A \opa{get}{i} is similar to \op{GetEnqueue}, taking $O(\log n + \log^2p)$ steps when the vector has $n$ elements.  
%An \opa{index}{e} is similar to \op{IndexDequeue} (except operating on enqueues instead of dequeues) and would take $O(\log p)$ steps.
\here{doublecheck all preceding claims for accuracy.}
Building on the work described in this paper, Asbell and Ruppert \cite{AR23} have
designed a doubly-ended queue with polylogarithmic amortized step
complexity, which also yields a stack as a special case.  
This required a substantially different representation of the data stored in the ordering tree.
\here{?  They also showed that a randomized implementation of the ordering tree technique can
be used to obtain a (doubly-ended) queue with O(\here{?}) amortized expected step complexity,
beating the  $\Omega(\min(c,\log\log p))$ lower bound on deterministic algorithms.}
Whether the ordering tree could also be used to obtain a priority queue with polylogarithmic step complexity
remains an open question.

>>>>

We believe the approach used here to implement a lock-free queue 
could be applied to obtain other lock-free
data structures with a polylogarithmic step complexity.
For example, we can easily adapt our routines to implement a  vector data structure that stores a sequence and
provides three operations: \opa{Append}{e} to add an element \var{e} to the end of the sequence,
\opa{Get}{i} to read the \var{i}th element in the sequence, and
\opa{Index}{e} to compute the position of element \var{e} in the sequence.
%\here{Can omit rest of this paragraph to save space}
%An \opa{append}{e} is implemented like \opa{Enqueue}{e} in $O(\log p)$ steps.  
%A \opa{get}{i} is similar to \op{GetEnqueue}, taking $O(\log n + \log^2p)$ steps when the vector has $n$ elements.  
%An \opa{index}{e} is similar to \op{IndexDequeue} (except operating on enqueues instead of dequeues) and would take $O(\log p)$ steps.
\here{doublecheck all preceding claims for accuracy.}
Building on the work described in this paper, Asbell and Ruppert \cite{AR23} have
designed a doubly-ended queue with polylogarithmic amortized step
complexity, which also yields a stack as a special case.  
This required a substantially different representation of the data stored in the ordering tree.
Whether the ordering tree could also be used to obtain a priority queue with polylogarithmic step complexity
remains an open question.
